---
title: "Coursera Practical Machine Learning Final Project"
author: "Tim Andrews"
date: "September 18, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary

This is my submission for the final project for Coursera's Practical Machine Learning.  The purpose of the project is to predict a classe variable for twenty supplied samples, using machine learning techniques.

The R code below downloads supplied training and test data sets, and then runs a prediction based on a random forest model.  As there were over 19,000 observations in the supplied data set, due to PC limitations, I chose to work with just 1002 of them for training, and took another 331 to test my prediction for cross validation purposes.

The results of the random forest produces an error rate of 5.79%.  Satisfied  with that, I then ran a prediction against the twenty supplied samples.

```{r code}
require(caret)
require(ggplot2)

set.seed(8675309)

###Read in supplied training aned test sets
training <- read.csv(url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"), header=TRUE)
testing <- read.csv(url("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"), header=TRUE)

featurePlot(x=training[,c("num_window", "roll_belt", "pitch_belt")], y = training$classe, plot="pairs")

###19K obervations are too many for my machine to handle, need to trim it down to 1333 (1002 train, 331 initial testing)
newTraining <- training[sample(nrow(training), 1333),]
inTrain <- createDataPartition(newTraining$classe, p = 3/4)[[1]]
usableTraining <- newTraining[inTrain,]
usableTesting <- newTraining[-inTrain,]

##create a random forest fit
fit <- train(classe ~ 
             +num_window
             +roll_belt
             +pitch_belt
             +yaw_belt
             +total_accel_belt
             +gyros_belt_x
             +gyros_arm_x
             +roll_dumbbell
             +total_accel_dumbbell
             +gyros_dumbbell_x
             +gyros_forearm_x
             +gyros_belt_y
             +gyros_arm_y
             +pitch_dumbbell
             +gyros_belt_z
             +gyros_arm_z
             +yaw_dumbbell
             +accel_belt_x
             +accel_arm_x
             +accel_belt_y
             +accel_arm_y
             +accel_belt_z
             +accel_arm_z
             +magnet_belt_x
             +magnet_arm_x
             +magnet_belt_y
             +magnet_arm_y
             +magnet_belt_z
             +magnet_arm_z
             +roll_arm
             +pitch_arm
             +yaw_arm
             +total_accel_arm
             +gyros_dumbbell_y
             +gyros_forearm_y
             +gyros_dumbbell_z
             +gyros_forearm_z
             +accel_dumbbell_x
             +accel_forearm_x
             +accel_dumbbell_y
             +accel_forearm_y
             +accel_dumbbell_z
             +accel_forearm_z
             +magnet_dumbbell_x
             +magnet_forearm_x
             +magnet_dumbbell_y
             +magnet_forearm_y
             +magnet_dumbbell_z
             +magnet_forearm_z
             +roll_forearm
             +pitch_forearm
             
             , method="rf", prox=TRUE, data=usableTraining)

###estimated error rate
fit$finalModel
#OOB estimate of  error rate: 5.79%

###create a prediction
usablePrediction <- predict(fit, newdata = usableTesting)

###validate results with a confusion matrix
confusionMatrix(usableTesting$classe, usablePrediction)
#Accuracy : 0.9366          
#95% CI : (0.9047, 0.9603)
#No Information Rate : 0.287          
#P-Value [Acc > NIR] : < 2.2e-16 


###predict on final, supplied testing set
finalPrediction <- predict(fit, newdata = testing)
#print(finalPrediction)
```
